Constraints on quasi--normal-mode frequencies with LIGO-Virgo binary--black-hole observations

### 2021-05-24 papers/intro_paper/intro_paper.tex
edits following Alessandra's comments

%Ground-based detectors, such as LIGO and Virgo, are most sensitive to stellar-mass BH binaries that form remnants with mass $\gtrsim 50 M_{\odot}$ and merge near the minimum of their sensitivity band ($\sim 100$ Hz). As a consequence, the remnant object \textit{rings down} at frequencies where the sensitivity is worse, leaving very little signal-to-noise ratio (SNR) in the post-merger signal. 
\textcolor{blue}{With current ground-based detectors, only a small fraction of BBH coalescences lead to a detectable post-merger GW signal and, even for these events, the signal-to-noise ratio (SNR) in the post-merger signal is, in most cases, only slightly above threshold for detection~\cite{LIGOScientific:2019fpa,Abbott:2020jks}. This is likely to remain the case in the near future~\cite{Berti:2016lat}.}
%
\textcolor{blue}{This is particularly important, given that} most of the tests mentioned above restrict the data to this post-merger signal which, while being agnostic to deviations in the pre-merger dynamics, significantly reduces parameter-estimation capabilities. Added to a lack of SNR is the ambiguity in clearly defining a start time for the ringdown phase (discussed, for example, in Refs.~\cite{Berti:2007fi,Baibhav:2017jhs,Bhagwat:2017tkm}). \textcolor{blue}{Defining such a time is particularly important in order to avoid possible systematic errors that may occur if one tries to apply perturbation theory in a regime where non-linearities are still important}. In some tests~\cite{Carullo:2018gah,Carullo:2019flw}, the ringdown start time has been left as a free parameter to be estimated directly from the data. Whereas, in other cases, the ringdown start time is predicted from corresponding full-signal parameter-estimation analyses (see \texttt{pyRing} analysis in Ref.~\cite{Abbott:2020jks}). Uncertainties
in estimates of the ringdown start-time, as well as an overall lack of
SNR in the post-merger signal, given typical sensitivities of
ground-based detectors, can result in significant statistical uncertainties in the measurement of some of the QNM properties. This is especially true for the measurement of the QNM damping times which are, in general, harder to measure from the data. Hence, one might want to look at alternate methods to measure QNMs using information from as much as the signal as possible.

@@

%%%%%%%%%%%%%%%%%
\section{Constraints on QNM frequencies using LIGO-Virgo data}
\label{sec:results_lvc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% O1-O2 events
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
        \includegraphics[width=0.5\textwidth]{figures/rin_pseob_results_v2.pdf}\includegraphics[width=0.3\textwidth]{figures/rin_all_events_bounds.pdf}
        \caption{\emph{Left panel}:The 90\% credible levels of the posterior probability distribution of the fractional deviations in the frequency and damping time of the $(2, 2)$ QNM, $(\df{220},\dtau{220})$ and their corresponding one-dimensional marginalized posterior distributions, for events from O1, O2 and O3a passing a SNR threshold of $8$ in both the pre- and post-merger signal. The solid purple curve marks the best single-event constraint, GW150914, whereas the contraints from the other events are indicated by the dash-dot curves. The joint constraints on $(\df{220},\dtau{220})$ obtained multiplying the likelihoods from individual events is given by the filled grey contours, while the hierarchical method of combination yields the black dot dashed curves. \emph{Right panel}: 90\% credible interval on the one-dimensional marginalised posteriors on $\delta \sigma_i=(\df{220},\dtau{220})$, colored by the median redshifted total mass $(1 + z)M$, inferred assuming GR. Filled gray (unfilled black) triangles mark the constraints obtained when all the events are combined by multiplying likelihoods (hierarchically). The purple square marker indicates constraints from the single most-prominent event, GW150914.}
        \label{fig:o1o2_events}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The LIGO-Virgo Collaboration recently released their testing GR
catalogue containing results for events observed
during O3a~\cite{Abbott:2020jks}. For the test that we here present, the results shown in Ref.~\cite{Abbott:2020jks} only include events which pass a threshold for the median detector-frame total mass $\geq 90 \Mo$ and SNRs in the pre- and
post-merger regions $\geq 8$~\footnote{The pre- and post-merger regions of the signal are identified from the signal's power before and after the signal reaches the peak's amplitude, which is determined from the maximum of the likelihood function obtained with the parameter-estimation analysis. The SNR values are listed in Table IV of Ref.~\cite{Abbott:2020jks}}. 
%
The SNR threshold ensures that the signal contains sufficient
information in both the inspiral and merger stages to break the
degeneracy between the binary's total mass and the non-GR deviations
$(\df{220}, \dtau{220})$. Such strong degeneracy \textcolor{blue}{, which arises from the fact that the frequency and damping time inverse are inversely proportional to the total mass~\cite{Berti:2005ys} such that one can always obtain the same frequency or damping time by appropriately increasing (or decreasing) the total mass and the deviation parameters in the same direction,} is present for
low-SNR events with negligible higher-modes and for which only the
post-merger is detectable rendering the measurement of $(\df{220},
\dtau{220})$ impossible for those cases (see Appendix~\ref{sec:correlation} for more details). On the other hand the
total mass threshold $\geq 90 \Mo$ was employed due to the fact that
this analysis is computationally expensive, and also because we
expected these events to be the most promising for ringdown
studies. However, since the SNR threshold alone should be sufficient
for the analysis, for this paper we run the test on all the events
listed in Ref.~\cite{Abbott:2020jks} that have SNRs in the pre- and
post-merger regions $\geq 8$, without imposing any mass threshold.

@@

However, in most non-GR theories, the deviations parameters $(\df{220},\dtau{220})$ depend in general on the source's parameters, so if GR were to be wrong, one should expect their value to vary across the GW signals observed by LIGO and Virgo.
As described in Ref.~\cite{Abbott:2020jks}, we can relax the assumption of a constant deviation across all events by using the hierarchical inference technique originally proposed in Refs.~\cite{Zimmerman:2019wzo,Isi:2019asy}. The general idea behind this technique is to assume that the non-GR parameters $(\df{220},\dtau{220})$ are drawn from a common underlying distribution, whose properties can be inferred from the population of events. Following~\cite{Zimmerman:2019wzo,Isi:2019asy,Abbott:2020jks} we model the population distribution with a Gaussian  $\mathcal{N}(\mu,\sigma)$ of unknown mean $\mu$ and standard deviation $\sigma$. Under those assumptions, the goal is then to directly measure a posterior distribution $P(\mu, \sigma |  \{d_j\}) $ for $\mu$ and $\sigma$ from a joint analysis of all the GW events. If GR is correct, then this posterior should be consistent with $\mu=0$ and $\sigma=0$. From Bayes' theorem it follows that~\cite{Isi:2019asy}:
%
\begin{equation}
P(\mu, \sigma |  \{d_j\}) \propto P(\mu,\sigma)\prod _{j=1}^N P(d_j |\, \mu, \sigma) \,,
\end{equation}
%
where $P(\mu,\sigma)$ is the prior (also known as hyperprior) on ($\mu,\sigma$), and $P(d_j | \mu, \sigma) $ can be written in terms of the individual likelihoods of a given non-GR parameter $\xi_{\text{nGR}}$ using~\cite{Isi:2019asy}
%
\begin{equation}
P(d_j | \mu, \sigma) = \int P(d_j |\, \xi_{\text{nGR}})  P(\xi_{\text{nGR}} |\, \mu, \sigma) \,d\xi_{\text{nGR}} \,.
\end{equation}
%
Here $P(\xi_{\text{nGR}} | \,\mu, \sigma) = \mathcal{N}(\mu,\sigma)$ by construction and $P(d_j |\, \xi_{\text{nGR}})$ is the likelihood for the parameter $\xi_{\text{nGR}}$ for a given event $d_j$ that is computed from the standard parameter estimation analysis. After obtaining $P(\mu, \sigma |  \{d_j\})$, we can then infer a population distribution for a given non-GR parameter $\xi_{\text{nGR}}$ using~\cite{Isi:2019asy}:
%
\begin{equation}\label{hier_combine}
P(\xi_{\text{nGR}} | \{d_j\}) = \int P(\xi_{\text{nGR}} |\, \mu, \sigma)\,P(\mu, \sigma |  \{d_j\})\,d\mu\,d\sigma \,.
\end{equation}
%
Notice that, if we fix $\sigma=0$, this approach is equivalent to assume that all events share the same non-GR parameter $\xi_{\text{nGR}} =\mu$ and Eq.~\eqref{hier_combine} reduces to the joint likelihood~\cite{Zimmerman:2019wzo}. In practice, we use the \texttt{stan}-based code~\cite{stan} developed and used in Refs.~\cite{Isi:2019asy,Abbott:2020jks} to obtain $P(\mu, \sigma |  \{d_j\}) $ and compute $P(\xi_{\text{nGR}} | \{d_j\})$.

The 1D posteriors for $\df{220}$ and $ \dtau{220}$ obtained with this technique are shown in Fig.~\ref{fig:o1o2_events} (dash-dotted curves) with corresponding median and 90\% credible interval given by: 
%
\begin{equation}
\df{220}=0.03^{+0.10}_{-0.09}\,,\quad \dtau{220}=0.10^{+0.44}_{-0.39}\,.
\end{equation}
Compared to Ref.~\cite{Abbott:2020jks} these constraints are almost a factor $\sim 4$ more stringent for $\df{220}$ and a factor $\sim 2$ for $\dtau{220}$. Similar improvements hold for the hyperparameters: $\df{220}\,(\mu=0.03^{+0.06}_{-0.05},\,\sigma<0.09)$ and $\dtau{220}\,(\mu=0.11^{+0.21}_{-0.19},\,\sigma<0.39)$.

>>>>

%Ground-based detectors, such as LIGO and Virgo, are most sensitive to stellar-mass BH binaries that form remnants with mass $\gtrsim 50 M_{\odot}$ and merge near the minimum of their sensitivity band ($\sim 100$ Hz). As a consequence, the remnant object \textit{rings down} at frequencies where the sensitivity is worse, leaving very little signal-to-noise ratio (SNR) in the post-merger signal. 
\textcolor{blue}{With current ground-based detectors, only a small fraction of BBH coalescences lead to a detectable post-merger GW signal and, even for these events, the signal-to-noise ratio (SNR) in the post-merger signal is, in most cases, only slightly above threshold for detection~\cite{LIGOScientific:2019fpa,Abbott:2020jks}.}
%
\textcolor{blue}{This is particularly important, given that} most of the tests mentioned above restrict the data to this post-merger signal which, while being agnostic to deviations in the pre-merger dynamics, significantly reduces parameter-estimation capabilities. Added to a lack of SNR is the ambiguity in clearly defining a start time for the ringdown phase (discussed, for example, in Refs.~\cite{Berti:2007fi,Baibhav:2017jhs,Bhagwat:2017tkm}). \textcolor{blue}{Defining such a time is particularly important in order to avoid possible systematic errors that may occur if one tries to extract the fundamental QNMs too early after merger, since the higher-order overtones dominate the ringdown signal closer to merger~\cite{Buonanno:2006ui,Giesler:2019uxc}}. In some tests~\cite{Carullo:2018gah,Carullo:2019flw}, the ringdown start time has been left as a free parameter to be estimated directly from the data. Whereas, in other cases, the ringdown start time is predicted from corresponding full-signal parameter-estimation analyses (see \texttt{pyRing} analysis in Ref.~\cite{Abbott:2020jks}). Uncertainties
in estimates of the ringdown start-time, as well as an overall lack of
SNR in the post-merger signal, given typical sensitivities of
ground-based detectors, can result in significant statistical uncertainties in the measurement of some of the QNM properties. This is especially true for the measurement of the QNM damping times which are, in general, harder to measure from the data. Hence, one might want to look at alternate methods to measure QNMs using information from as much as the signal as possible.

@@

%%%%%%%%%%%%%%%%%
\section{Constraints on QNM frequencies using LIGO-Virgo data}
\label{sec:results_lvc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% O1-O2 events
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
        \includegraphics[width=0.5\textwidth]{figures/rin_pseob_results_v2.pdf}\includegraphics[width=0.3\textwidth]{figures/rin_all_events_bounds.pdf}
        \caption{\emph{Left panel}:The 90\% credible levels of the posterior probability distribution of the fractional deviations in the frequency and damping time of the $(2, 2)$ QNM, $(\df{220},\dtau{220})$ and their corresponding one-dimensional marginalized posterior distributions, for events from O1, O2 and O3a passing a SNR threshold of $8$ in both the pre- and post-merger signal. The solid purple curve marks the best single-event constraint, GW150914, whereas the contraints from the other events are indicated by the dash-dot curves. The joint constraints on $(\df{220},\dtau{220})$ obtained multiplying the likelihoods from individual events is given by the filled grey contours, while the hierarchical method of combination yields the black dot dashed curves (\textcolor{blue}{only shown in the 1D posteriors plots}). \emph{Right panel}: 90\% credible interval on the one-dimensional marginalised posteriors on $\delta \sigma_i=(\df{220},\dtau{220})$, colored by the median redshifted total mass $(1 + z)M$, inferred assuming GR. Filled gray (unfilled black) triangles mark the constraints obtained when all the events are combined by multiplying likelihoods (hierarchically). The purple square marker indicates constraints from the single most-prominent event, GW150914.}
        \label{fig:o1o2_events}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The LIGO-Virgo Collaboration recently released their testing GR
catalogue containing results for events observed
during O3a~\cite{Abbott:2020jks}. For the test that we here present, the results shown in Ref.~\cite{Abbott:2020jks} only include events which pass a threshold for the median detector-frame total mass $\geq 90 \Mo$ and SNRs in the pre- and
post-merger regions $\geq 8$~\footnote{The pre- and post-merger regions of the signal are identified from the signal's power before and after the signal reaches the peak's amplitude, which is determined from the maximum of the likelihood function obtained with the parameter-estimation analysis. The SNR values are listed in Table IV of Ref.~\cite{Abbott:2020jks}}. 
%
The SNR threshold ensures that the signal contains sufficient
information in both the inspiral and merger stages to break the
degeneracy between the binary's total mass and the non-GR deviations
$(\df{220}, \dtau{220})$. Such strong degeneracy \textcolor{blue}{arises from the fact that the complex QNM frequencies are inversely proportional to the total mass~\cite{Berti:2005ys} such that one can always obtain the same fundamental frequency or damping time by appropriately increasing (or decreasing) the total mass and the deviation parameters in the same direction. This degeneracy is especially important} in
low-SNR events with negligible higher-modes and for which only the
post-merger is detectable, rendering the measurement of $(\df{220},
\dtau{220})$ impossible for those cases  (see Appendix~\ref{sec:correlation} for more details). On the other hand the
total mass threshold $\geq 90 \Mo$ was employed due to the fact that
this analysis is computationally expensive, and also because we
expected these events to be the most promising for ringdown
studies. However, since the SNR threshold alone should be sufficient
for the analysis, for this paper we run the test on all the events
listed in Ref.~\cite{Abbott:2020jks} that have SNRs in the pre- and
post-merger regions $\geq 8$, without imposing any mass threshold.

@@

However, in most non-GR theories, the deviations parameters $(\df{220},\dtau{220})$ depend in general on the source's parameters, so if GR were to be wrong, one should expect their value to vary across the GW signals observed by LIGO and Virgo.
As described in Ref.~\cite{Abbott:2020jks}, we can relax the assumption of a constant deviation across all events by using the hierarchical inference technique originally proposed in Refs.~\cite{Zimmerman:2019wzo,Isi:2019asy}. The general idea behind this technique is to assume that the non-GR parameters $(\df{220},\dtau{220})$ are drawn from a common underlying distribution, whose properties can be inferred from the population of events. Following~\cite{Zimmerman:2019wzo,Isi:2019asy,Abbott:2020jks} we model the population distribution with a Gaussian  $\mathcal{N}(\mu,\sigma)$ of unknown mean $\mu$ and standard deviation $\sigma$. Under those assumptions, the goal is then to directly measure a posterior distribution $P(\mu, \sigma |  \{d_j\}) $ for $\mu$ and $\sigma$ from a joint analysis of all the GW events. If GR is correct, then this posterior should be consistent with $\mu=0$ and $\sigma=0$. From Bayes' theorem it follows that~\cite{Isi:2019asy}:
%
\begin{equation}
P(\mu, \sigma |  \{d_j\}) \propto P(\mu,\sigma)\prod _{j=1}^N P(d_j |\, \mu, \sigma) \,,
\end{equation}
%
where $P(\mu,\sigma)$ is the prior (also known as hyperprior) on ($\mu,\sigma$), and $P(d_j | \mu, \sigma) $ can be written in terms of the individual likelihoods of a given non-GR parameter $\xi_{\text{nGR}}$ using~\cite{Isi:2019asy}
%
\begin{equation}
P(d_j | \mu, \sigma) = \int P(d_j |\, \xi_{\text{nGR}})  P(\xi_{\text{nGR}} |\, \mu, \sigma) \,d\xi_{\text{nGR}} \,.
\end{equation}
%
Here $P(\xi_{\text{nGR}} | \,\mu, \sigma) = \mathcal{N}(\mu,\sigma)$ by construction and $P(d_j |\, \xi_{\text{nGR}})$ is the likelihood for the parameter $\xi_{\text{nGR}}$ for a given event $d_j$ that is computed from the standard parameter estimation analysis. After obtaining $P(\mu, \sigma |  \{d_j\})$, we can then infer a population distribution for a given non-GR parameter $\xi_{\text{nGR}}$ using~\cite{Isi:2019asy}:
%
\begin{equation}\label{hier_combine}
P(\xi_{\text{nGR}} | \{d_j\}) = \int P(\xi_{\text{nGR}} |\, \mu, \sigma)\,P(\mu, \sigma |  \{d_j\})\,d\mu\,d\sigma \,.
\end{equation}
%
Notice that, if we fix $\sigma=0$, this approach is equivalent to assume that all events share the same non-GR parameter $\xi_{\text{nGR}} =\mu$ and Eq.~\eqref{hier_combine} reduces to the joint likelihood~\cite{Zimmerman:2019wzo}. In practice, we use the \texttt{stan}-based code~\cite{stan} developed and used in Refs.~\cite{Isi:2019asy,Abbott:2020jks} to obtain $P(\mu, \sigma |  \{d_j\}) $ and compute $P(\xi_{\text{nGR}} | \{d_j\})$. \textcolor{blue}{We note that the current implementation of this analysis, as originally defined in~\cite{Isi:2019asy}, is only defined for 1D posteriors, therefore below we only show 1D posteriors for the hierarchical analysis.}

The posteriors for $\df{220}$ and $ \dtau{220}$ obtained with this technique are shown in Fig.~\ref{fig:o1o2_events} (dash-dotted curves) with corresponding median and 90\% credible interval given by: 
%
\begin{equation}
\df{220}=0.03^{+0.10}_{-0.09}\,,\quad \dtau{220}=0.10^{+0.44}_{-0.39}\,.
\end{equation}
Compared to Ref.~\cite{Abbott:2020jks} these constraints are almost a factor $\sim 4$ more stringent for $\df{220}$ and a factor $\sim 2$ for $\dtau{220}$. Similar improvements hold for the hyperparameters: $\df{220}\,(\mu=0.03^{+0.06}_{-0.05},\,\sigma<0.09)$ and $\dtau{220}\,(\mu=0.11^{+0.21}_{-0.19},\,\sigma<0.39)$.
